---
title: "Untitled"
author: "Shelly Trigg"
date: "4/18/2026"
output: html_document
editor_options: 
  chunk_output_type: console
---
This is an analysis of the tables resulting from differentialAbundance run on 4/18/26 by Shelly Wanamaker of 217 samples using the matrix file https://gannet.fish.washington.edu/metacarcinus/USDA_MetaOmics/Cvirg_RNAseq/20260418_diffabund/all_merged_gene_countsCorrected2_length_scaled.tsv, the samplesheet https://gannet.fish.washington.edu/metacarcinus/USDA_MetaOmics/Cvirg_RNAseq/20260418_diffabund/ and contrasts file https://gannet.fish.washington.edu/metacarcinus/USDA_MetaOmics/Cvirg_RNAseq/20260418_diffabund/rnaseq_diffabundance_contrasts.csv. It also includes tables resulting from post-analysis data integration run on the same day. 

load libraries
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(purrr)
library(stringr)
library(UpSetR)
library(tibble)
library(ComplexHeatmap)
library(circlize)

```



Prepare merged matrix of gene counts scaled by length 
```{r}
oldallds <- read.csv("../../data/rnaseq_gene_counts/all_merged_gene_counts_length_scaled.tsv", sep = "\t")

ds1 <- read.csv("../../data/rnaseq_gene_counts/salmon.merged.gene_counts_length_scaled_dataset1.tsv", sep = "\t")
ds2 <- read.csv("../../data/rnaseq_gene_counts/salmon.merged.gene_counts_length_scaled_dataset2.tsv", sep = "\t")
ds3 <- read.csv("../../data/rnaseq_gene_counts/salmon.merged.gene_counts_length_scaled_dataset3.tsv", sep = "\t")
ds4 <- read.csv("../../data/rnaseq_gene_counts/salmon.merged.gene_counts_length_scaled_dataset4.tsv", sep = "\t")
ds5 <- read.csv("../../data/rnaseq_gene_counts/salmon.merged.gene_counts_length_scaled_dataset5.tsv", sep = "\t")


allds <- full_join(ds1,ds2, by = c("gene_name", "gene_id"))
allds <- full_join(allds,ds3, by = c("gene_name", "gene_id"))
allds <- full_join(allds,ds4, by = c("gene_name", "gene_id"))
allds <- full_join(allds,ds5, by = c("gene_name", "gene_id"))
allds$gene_name <- NULL


#confirm if old and new tables are using the correct data
View(merge(allds %>% pivot_longer(cols = 2:ncol(allds),names_to = "samples", values_to = "VSTcounts"), oldallds%>% pivot_longer(cols = 2:ncol(oldallds),names_to = "samples", values_to = "VSTcounts"), by = c("gene_id", "samples")))

#filter for dataset2 sample (e.g. SRX7656987)
#all values are the same so data is correct

#this is the file that was linked above
write.table(allds, "../../data/rnaseq_gene_counts/all_merged_gene_countsCorrected2_length_scaled.tsv", sep = "\t", row.names = F, quote = F)


```

read in data
```{r}
ds1_filt <- read.csv("../../analyses/20260418_diffAbund_allData/ds1_condition_sensitive_tolerant.deseq2.results_filtered.tsv", sep = "\t")

ds2_filt <- read.csv("../../analyses/20260418_diffAbund_allData/ds2_condition_sensitive_tolerant.deseq2.results_filtered.tsv", sep = "\t")

ds3_filt <- read.csv("../../analyses/20260418_diffAbund_allData/ds3_condition_sensitive_tolerant.deseq2.results_filtered.tsv", sep = "\t")

ds4_filt <- read.csv("../../analyses/20260418_diffAbund_allData/ds4_condition_sensitive_tolerant.deseq2.results_filtered.tsv", sep = "\t")

ds5_filt <- read.csv("../../analyses/20260418_diffAbund_allData/ds5_condition_sensitive_tolerant.deseq2.results_filtered.tsv", sep = "\t")

allds_filt <- read.csv("../../analyses/20260418_diffAbund_allData/allds_condition_sensitive_tolerant.deseq2.results_filtered.tsv", sep = "\t")


#from https://gannet.fish.washington.edu/metacarcinus/USDA_MetaOmics/Cvirg_RNAseq/20260418_diffabund/allds/tables/processed_abundance/

norm_counts0418<-read.csv("../../analyses/20260418_diffAbund_allData/all.normalised_counts.tsv", sep ="\t")

vst_counts0418 <- read.csv("../../analyses/20260418_diffAbund_allData/all.vst.tsv", sep ="\t")

samplesheet <- read.csv("../../data/differential_abundance_sheets/rnaseq_difabundance_fullset_s3_samplesheet.csv")
```

add dataset column to samplesheet
```{r}
samplesheet$study <- (c(rep("study_3",10), rep("study_1",61),rep("study_5",62), rep("study_2",40), rep("study_4",44)))
```

plot gene abundances of DEGs from all data together
```{r}
jpeg("../20260418_diffAbund_allData/Boxplots_VST_18DEGs_DESeq2.jpeg", width = 7, height = 8, units = "in", res = 300)
data.frame(vst_counts0418) %>% pivot_longer(cols = 2:218, names_to = "sample", values_to = "VST") %>% filter(gene_id %in% allds_filt$gene_id) %>% right_join(samplesheet %>% dplyr::select(sample,study, condition)) %>% ggplot() + geom_boxplot(aes(x = condition, y = VST), fill = NA, outlier.shape = NA) + geom_jitter(aes(x=condition,y=VST, color = study), shape = 16, size = 0.7, alpha = 0.7) +  facet_wrap(~gene_id, scales = "free")+ theme_bw()+ theme(legend.position = "top", axis.text.x =element_text(angle = 20))
dev.off()
```

plot pca
```{r}
#prep data for PCA
#create matrix from counts matrix
vst_counts0418_filt <- vst_counts0418 %>% filter(gene_id %in% allds_filt$gene_id)
rownames(vst_counts0418_filt)<- vst_counts0418_filt[,1]
vst_counts0418_filt <- t(vst_counts0418_filt)
vst_counts0418_filt <- vst_counts0418_filt[-1,]
vst_counts0418_m <- apply(vst_counts0418_filt,2,as.numeric)
rownames(vst_counts0418_m)<- rownames(vst_counts0418_filt)

pca_filt <- prcomp(vst_counts0418_m, center = T, scale = T)

pca_filt_summ <- summary(pca_filt)

pca_filt_meta <- merge(samplesheet[,c("sample", "condition","study")],cbind(sample =rownames(pca_filt$x), pca_filt$x), by = "sample") %>% mutate(across(4:ncol(.), as.numeric))


jpeg("../20260418_diffAbund_allData/PCA_VST_18DEGs_DESeq2.jpeg", width = 4, height = 3, units = "in", res = 300)
ggplot(pca_filt_meta, aes(PC1, PC2,color = study, shape = condition))+ geom_point() + theme_bw() + labs(y = paste("PC2"," (",formatC(pca_filt_summ$importance[5] * 100,digits=2,format="f"),"%)", sep = ""), x = paste("PC1"," (",formatC(pca_filt_summ$importance[2] * 100,digits=2,format="f"),"%)", sep = "")) + ggtitle("PCA of VST gene expression for 18 genes")+ scale_shape_manual(values = c(4,16))
dev.off()

```

merging data from study specific diffabund
```{r}

# Automatically grab all *_filt objects
obj_names <- ls(pattern = "ds[0-9]_filt$")
#obj_names <- obj_names[which(obj_names !="ds2_filt")]
#obj_names <- c(obj_names,"allds_filt")
df_list <- mget(obj_names)

# Rename columns using object name as study prefix
df_list_renamed <- imap(df_list, function(df, obj_name) {

  study <- str_remove(obj_name, "_filt$")  # ds1, ds2, etc.

  df %>%
    rename_with(
      ~ paste0(study, "_", .x),
      .cols = -gene_id
    )
})

# Full join across all studies
merged_df <- reduce(df_list_renamed, full_join, by = "gene_id")


# Identify presence via baseMean columns
baseMean_cols <- grep("_baseMean$", names(merged_df), value = TRUE)

merged_df <- merged_df %>%
  mutate(
    studies_detected_in = rowSums(!is.na(across(all_of(baseMean_cols))))
  ) %>%
  arrange(desc(studies_detected_in))

#write table of merged results
write.table(merged_df, "ds1-5_allDEGs_merged.csv", sep = ",", row.names = F, quote = F)

merged_df %>%
  ggplot(aes(x = studies_detected_in)) +
  geom_bar() +
  theme_minimal() +
  xlab("Number of studies gene appears in") +
  ylab("Number of genes")



# build binary membership dataframe
presence_df <- merged_df %>%
  select(all_of(baseMean_cols)) %>%
  mutate(across(
    everything(),
    ~ ifelse(is.na(.x), 0, 1)
  ))


# clean study names
colnames(presence_df) <- str_remove(colnames(presence_df), "_baseMean$")

# convert to plain data.frame
presence_df <- as.data.frame(presence_df)


# make upset plot
jpeg("../20260418_diffAbund_allData/Upset_5studies.jpeg", width = 5, height = 3, units = "in", res = 300)
upset(
  as.data.frame(presence_df),
  sets = colnames(presence_df),
  order.by = "freq",
  keep.order = TRUE
)
dev.off()

```


```{r}
lfc_cols <- grep("_log2FoldChange$", names(merged_df), value = TRUE)

consistent_dir <- merged_df %>%
  mutate(
    # matrix of LFCs
    mat = select(., all_of(lfc_cols)) %>% as.matrix(),

    # convert to sign matrix: -1, 0, 1, NA
    sign_mat = sign(mat),

    # count only non-NA values
    n_observed = rowSums(!is.na(sign_mat)),

    # count direction occurrences
    n_up = rowSums(sign_mat == 1, na.rm = TRUE),
    n_down = rowSums(sign_mat == -1, na.rm = TRUE),

    direction = case_when(
      n_up > 0 & n_down == 0 ~ "up",
      n_down > 0 & n_up == 0 ~ "down",
      n_up > 0 & n_down > 0 ~ "mixed",
      TRUE ~ NA_character_
    ),

    consistency = pmax(n_up, n_down) / (n_up + n_down)
  )  %>%
  filter(n_observed >= 2, direction != "mixed")
```

make heatmap of logfc across studies
```{r}
#filt_cons_dir <- consistent_dir %>% dplyr::select(gene_id, ds1_log2FoldChange,ds2_log2FoldChange,ds3_log2FoldChange,ds4_log2FoldChange,ds5_log2FoldChange) %>% column_to_rownames(var="gene_id") %>% rename_with(~gsub("_.*", "", .)) %>% as.matrix() 
#filt_cons_dir[is.na(filt_cons_dir)] <- 0

#filter for consistently upregulated genes

filt_cons_dir_pos <- consistent_dir %>% filter(direction == "up")%>% dplyr::select(gene_id, ds1_log2FoldChange,ds2_log2FoldChange,ds3_log2FoldChange,ds4_log2FoldChange,ds5_log2FoldChange) %>% column_to_rownames(var="gene_id") %>% rename_with(~gsub("_.*", "", .)) %>% as.matrix() 

filt_cons_dir_pos[is.na(filt_cons_dir_pos)] <- 0

rngmax <- max(filt_cons_dir_pos, na.rm = TRUE)

rngmin <- min(filt_cons_dir_pos, na.rm = TRUE)

#define color mapping
col_fun <- colorRamp2(
  c(0, 1, 8),
  c("white", "#FEE5D9", "#CB181D")
)

#plot heatmap of genes with consistent positive FC

jpeg("../20260418_diffAbund_allData/Heatmap_83upDEGs_5studies.jpeg", width = 3, height = 3, units = "in", res = 300)
Heatmap(
  filt_cons_dir_pos,
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE, show_row_names = FALSE, name = "log2 FC")
dev.off()

#filter for consistently downregulated genes
filt_cons_dir_neg <- consistent_dir %>% filter(direction == "down")%>% dplyr::select(gene_id, ds1_log2FoldChange,ds2_log2FoldChange,ds3_log2FoldChange,ds4_log2FoldChange,ds5_log2FoldChange) %>% column_to_rownames(var="gene_id") %>% rename_with(~gsub("_.*", "", .)) %>% as.matrix() 

filt_cons_dir_neg[is.na(filt_cons_dir_neg)] <- 0

rngmax <- max(filt_cons_dir_neg, na.rm = TRUE)
rngmin <- min(filt_cons_dir_neg, na.rm = TRUE)

#define color mapping
col_fun <- colorRamp2(
  c(-7, -1, 0),
  c("#08519C", "#DEEBF7", "white")
)

#plot heatmap
jpeg("../20260418_diffAbund_allData/Heatmap_67downDEGs.jpeg", width = 3, height = 3, units = "in", res = 300)
Heatmap(
  filt_cons_dir_neg,
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE, show_row_names = FALSE, name = "log2 FC")
dev.off()



```

Find genes in common between IDA and PADI methods
```{r}

vst_counts0418%>% filter(gene_id %in% allds_filt[which(allds_filt$PADIoverlap),"gene_id"]) %>% column_to_rownames(var="gene_id") %>% rename_with(~gsub("_.*", "", .)) %>% as.matrix() 
  
  
data.frame(vst_counts0418) %>% pivot_longer(cols = 2:218, names_to = "sample", values_to = "VST") %>% filter(gene_id %in% allds_filt[which(allds_filt$PADIoverlap),"gene_id"]) %>%   
 right_join(samplesheet %>% dplyr::select(sample,study, condition)) %>% ggplot() + geom_boxplot(aes(x = condition, y = VST), fill = NA, outlier.shape = NA) + geom_jitter(aes(x=condition,y=VST, color = study), shape = 16, size = 0.7, alpha = 0.7) +  facet_wrap(~gene_id, scales = "free")+ theme_bw()+ theme(legend.position = "top", axis.text.x =element_text(angle = 20))




```
